Cartogram Mapping and its Application to Cancer Data Visualisation

### 2019-09-09 review_article.Rmd
Figures using cowplots

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
library(ggplot2)
library(grid)
library(png)
library(gridExtra)
```
<!--

@@

```{r choropleth-grid-create, message=FALSE, fig.cap = "Public facing choropleth cancer maps. From right to left and top to bottom: [a. Globocan 2012: Estimated  Cancer Incidence, Mortality and Prevalence](http://globocan.iarc.fr/Pages/Map.aspx)  (Incidence rates per 100,000 for all cancers, male and female),  [b. Bowel Cancer Australia Atlas](http://www.bowelcanceratlas.org/), [c. Atlas of Cancer in QLD - Relative incidence ration, Lung cancer in males](https://cancerqld.org.au/research/queensland-cancer-statistics/queensland-cancer-atlas/), [d. CINA+ Online Cancer in North America](http://www.cancer-rates.info/naaccr/) (Incidence rates of all invasive cancers per 100,000, male and female), [e. Atlas of Childhood Cancer in Ontario](http://www.pogo.ca/research-data/pogo-atlas/) (Incidence rate of childhood cancer per 100,000, men and women), [f. An Atlas of Cancer in South Australia](https://www.cancersa.org.au/assets/images/pdfs/An%20Atlas%20of%20Cancer%20in%20South%20Australia%20-%20Full%20Report.pdf) (Incidence rate per 100,00, lung cancer in females), [g. The Environment and Health Atlas of England and Wales](http://www.envhealthatlas.co.uk/eha/Breast/) (Lung cancer in females), [h. Map of Cancer Mortality Rates in Spain](http://elpais.com/elpais/2014/10/06/media/1412612722_141933.html) (Relative risk, Lung cancer, men vs women), [i. United States Cancer Statistics: An Interactive Cancer Statistics Website](https://nccd.cdc.gov/DCPC_INCA/) (Incidence rate per 100,000, all cancer types, men and women"}

plot1 <- png::readPNG('cancer_map_images/choropleths/3.choropleth.png')
plot2 <- png::readPNG('cancer_map_images/choropleths/15.choropleth.png')
plot3 <- png::readPNG('cancer_map_images/choropleths/25.choropleth.png')
plot4 <- png::readPNG('cancer_map_images/choropleths/20.choropleth1.png')
plot5 <- png::readPNG('cancer_map_images/choropleths/26.choropleth.png')
plot6 <- png::readPNG('cancer_map_images/choropleths/14.choropleth2.png')
plot7 <- png::readPNG('cancer_map_images/choropleths/21.choropleth.png')
plot8 <- png::readPNG('cancer_map_images/choropleths/23.choropleth.png')
plot9 <- png::readPNG('cancer_map_images/choropleths/7.choropleth.png')

choropleth_grid <- gridExtra::grid.arrange(
                  grid::rasterGrob(plot1),
                  grid::rasterGrob(plot2), 
                  grid::rasterGrob(plot3),                
                  grid::rasterGrob(plot4), 
                  grid::rasterGrob(plot5), 
                  grid::rasterGrob(plot6),
                  grid::rasterGrob(plot7), 
                  grid::rasterGrob(plot8), 
                  grid::rasterGrob(plot9), ncol=3)

ggsave(filename = "figures/choropleth_grid.png", plot = choropleth_grid, device = "png", width = 6, height = 5, dpi = 300)
```

@@

While cancer maps are powerful visualisations that summarise complex statistical analyses, the statistics represented in these maps cannot tell the entire story. Supplementary graphs and plots are often included that add more depth and more information to the map. Maps can also utilise bar charts, tables and scatter-plots as useful additional displays. 
These displays can provide alternative views of the cancer statistics, as well as the supporting statistical outputs such as error, confidence intervals, distributions, sample/population sizes, standard deviation, as well as others.. 
When presenting cancer maps, @MACM believes the intuition derived from maps must be 'validated by rigorous statistical analyses.' These supplementary statistics aim to enable this validation. @CPISACA suggests additional materials such as 'good tables, graphs, and explanatory text' support understanding and inference derived from maps, ensuring the message communicated will be consistent across a range of viewers.

```{r supp-grid-create, fig.cap = "Interactive controls of displays in public facing choropleth cancer maps. From left to right and top to bottom: [a. Globocan 2018: Cancer Today](http://globocan.iarc.fr/Pages/Map.aspx)(Controls for indicator, sex, age groups, continents, and cancer types) , [b. Bowel Cancer Australia Atlas](http://www.bowelcanceratlas.org/) (Controls for indicator, age, year, and geographical areas), [c. Global Cancer Map](http://globalcancermap.com/) (Controls for indicator, and cancer type), [d. United States Cancer Statistics: Data Visualizations](https://gis.cdc.gov/Cancer/USCS/DataViz.html) (Controls for demographics, trends, geography, indicators, risk factors, sex, cancer type, and year)."}
i3 <- png::readPNG('cancer_map_images/interactivity/3.interactivity.png')
i15 <- png::readPNG('cancer_map_images/interactivity/15.interactivity.png')
i4 <- png::readPNG('cancer_map_images/interactivity/4.interactivity.png')
i5 <- png::readPNG('cancer_map_images/interactivity/5.interactivity.png')
i7 <- png::readPNG('cancer_map_images/interactivity/7.interactivity.png')

lay <- rbind(c(1,1,2,2,2,2),
             c(1,1,2,2,2,2),
             c(1,1,2,2,2,2),
             c(1,1,3,3,3,3),
             c(1,1,4,4,4,4),
             c(1,1,5,5,5,5))

supp_grid <- gridExtra::grid.arrange(
  grid::rasterGrob(i3), grid::rasterGrob(i15),
  grid::rasterGrob(i5), grid::rasterGrob(i7), grid::rasterGrob(i4), layout_matrix = lay)

ggsave(filename = "figures/supp_grid.png", plot = supp_grid, device = "png", width = 6, height = 5, dpi = 300)
```

**Interactivity**

The interactivity of modern maps enables supplementary information to be incorporated without cluttering the screen, such as in a tool tip feature. The use of these supports were found in a variety of online map sources. These interactive features provide additional information to help users understand and interpret the spatial distribution presented, as well as validate, explain or explore the presented statistics and their relationships to each other and/or their underlying spatial distribution. The interactive features of the publically avaialble maps identified by @roberts2019communication were.......

@@

### Socio-economic indicators
Socio-economic indicators can explain how the experience of cancer prevalence varies for various members of a society.
These indicators include unemployment rates, poverty rates and percentages, remoteness and education levels achieved though only a few atlases also explored the impact of rurality on cancer rates. 
The Human Development Index can be used to understand the socio-economic experience of a community, as can the Income levels as measured by World bank list of economies[@IARC_3]. 
The areas are usually ranked and allocated to quintiles to be presented as categories describing the ranking.

@@

<!-- There are many alternative views of the world-->
<!-- Tobler map projections-->
Choropleths may be considered true topological maps, however, if the land mass displayed covers enough of the globe, there must be a transformation or distortion to display the land in 2D. The amount of distortion is related to the distance covered by the landmass displayed @GAMP.
World map projections reflect the frequent perspectives used to view the earth. Choropleth maps will always be distorted if they cover enough of the globe, just like photographs of the globe from space. Choropleth creation requires choosing a map projection that shows a favourable distortion of the geography for presenting the set of spatial information. Selecting a display can prevent misinterpretation of global statistics, as global maps face the challenge of equitable displays of land mass on maps [@RSCW].
Maps that do not specify a projection can be considered to have some unknown projection. <!-- ??? cite tobler.-->
If the statistic presented on the map base relies on physical distance and is influenced by the topology there is no transformation needed, beyond choosing a reasonable projection.

@@

<!-- Why transform?-->
<!-- Common variables used to create cartograms, e.g. population, mortality -->
Cartograms provide an alternative visualisation method for statistical and geographical information.
The key difference between a choropleth and a cartogram is the desirable augmentation of the size, shape or distance of geographical areas [@ACTUC].
Cartograms may be seen as an extension of map transformations and projections. The favourable distortion is proportional to a value other other than actual earth size area [@NAC].
A disadvantage of the conventional map is that sparsely populated rural areas may be emphasized, whereas the areas representing cities are very small, making interpretation of spatial patterns very difficult.
The distortion of a cartogram accounts for the population density, preventing it from obscuring the spatial patterns [@TAAM]. 
The spatial transformation of map regions relative to the data emphasises the data distribution instead of land size [@CBATCC].
When visualising population statistics @ACTUC considers this equitable representation design 'more socially just', or honest [@NISCC], giving due attention to all members of the population and reducing the visual impact of large areas with small populations [@DMAHP]. 
@HEDP suggests that 'cancer occurs in people, not in geographical areas' and @CTTMB believe that spatial socio-economic data, like cancer rates, are best presented on a cartogram for urban areas as the population map base avoids allocating 'undue prominence' to rural areas. 
@MTMSIH encourage the use of cartograms to highlight small areas and uncover local-level inequalities.

@@

Square grids have been used by FiveThirtyEight, Bloomberg Business, The Guardian, The Washington Post, The New York Times and NPR. Where each area is represented by a square of the same dimensions.
Figure \ref{fig:ggtilegram} shows a square tilegram representing each state using a single square.

@@

Traditionally, epidemiologists as disease mappers would have understood the characteristics of a disease lifespan and movements. they would have also considered the interactions between geography, the environment and the people captured by a display.
It is important for map creators to develop and then test the message being communicated, epidemiologists are able to do this with their expert knowledge.
Static visualisations can help to aid the written results provided by experts, and pairing an explanation with a map or diagram can help to draw attention to a point, group or area of interest. They have the caveat of a need to provide enough information for the static displays to be understandable out of their original context.

Recent developments in technology have created an opportunity to migrate atlases from paper displays in books to interactive web atlases.
The medium chosen to communicate health data should always be driven by the message to communicate and the needs of the audience. This can be seen as a continuum, where the message is most important, static or animated graphics allw control over the display and interpretation. Where the needs of the audience is changeable and is the priority, the map creator can allow interactivity for map users to explore.

<!-- Animations -->
@CPISACA provide weather maps as well developed and appropriate examples of spatial displays to communicate to the general public and encourage those creating disease maps to leverage the success of weather displays.
The movement of a weather system will follow a forecasted path, the animation will communicate this as all map users can follow the path of the system across the geography over a specified period of time.

<!-- Interactivity-->
The move from static, print media displays toward interactivity allow users to develop maps to ask and answer their own questions.
Developmets and adoptions of this technology has allowed map users more control of the message displayed.
They are able to see the familiar shapes, zoom in to explore densely populated communities, or change the display completely to highlight the population space.
@CPISACA provide three questions that are generally when using a disease map:

@@

A single map display historically created by epidemiologists, may not effectively answer all three at the same time.
Especially when areas are small and hard to identify in a display.
However, with the ability to animate, map creators can specify and answer these questions by directing the attention of a map user to the areas of interest. Animations can be used to control the message being communicated by capturing and directing attention of users, it will also ensure all users follow the same interpretive path, as their eyes are drawn across the map to the same comparison points.

This oppportunity provides room to move away from the traditional approach of a single map for a single use, into a more explorative territory. Interactivity allows users to drive their own exploration, controlling a display through variable controls, links, tool tips, and data selection tools [@TGA].

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE, fig.width = 6, fig.height = 5)
library(ggplot2)
library(grid)
library(png)
library(gridExtra)
```
<!--

@@

```{r choropleth-grid-create, message=FALSE, fig.cap = "Public facing choropleth cancer maps.[a. The Environment and Health Atlas of England and Wales](http://www.envhealthatlas.co.uk/eha/Breast/) (Lung cancer in females)  (Incidence rates per 100,000 for all cancers, male and female),  [b. Bowel Cancer Australia Atlas](http://www.bowelcanceratlas.org/), [c. Atlas of Cancer in QLD - Relative incidence ration, Lung cancer in males](https://cancerqld.org.au/research/queensland-cancer-statistics/queensland-cancer-atlas/), [d. Globocan 2012: Estimated  Cancer Incidence, Mortality and Prevalence](http://globocan.iarc.fr/Pages/Map.aspx), [e. United States Cancer Statistics: An Interactive Cancer Statistics Website](https://nccd.cdc.gov/DCPC_INCA/) (Incidence rate per 100,000, all cancer types, men and women), [f. Map of Cancer Mortality Rates in Spain](http://elpais.com/elpais/2014/10/06/media/1412612722_141933.html) (Relative risk, Lung cancer, men vs women), [g. An Atlas of Cancer in South Australia](https://www.cancersa.org.au/assets/images/pdfs/An%20Atlas%20of%20Cancer%20in%20South%20Australia%20-%20Full%20Report.pdf) (Incidence rate per 100,000, lung cancer in females) "}
library(cowplot)

# Import the images
plot1 <- png::readPNG('cancer_map_images/choropleths/21.choropleth.png')
plot2 <- png::readPNG('cancer_map_images/choropleths/15.choropleth.png')
plot3 <- png::readPNG('cancer_map_images/choropleths/25.choropleth.png')
plot4 <- png::readPNG('cancer_map_images/choropleths/3.choropleth.png') 
plot5 <- png::readPNG('cancer_map_images/choropleths/7.choropleth.png')
plot6 <- png::readPNG('cancer_map_images/choropleths/23.choropleth.png')
plot7 <- png::readPNG('cancer_map_images/choropleths/26.choropleth.png')

ggdraw() +
  draw_plot(rasterGrob(plot1), 0, .75, 0.33, .25) +
  draw_plot(rasterGrob(plot2), 0.33, 0.75, 0.33, .25) +
  draw_plot(rasterGrob(plot3), 0.66, 0.5, .33, .5) +
  draw_plot(rasterGrob(plot4), 0, 0.5, .33, .25) +
  draw_plot(rasterGrob(plot5), 0.33, 0.5, .33, .25) +
  draw_plot(rasterGrob(plot6), 0, 0, .66, .5) +
  draw_plot(rasterGrob(plot7), .66, 0, .33, .5) +
  draw_plot_label(c("a", "b", "c", "d", "e", "f", "g"), 
    c(0.015, 0.33, 0.63, 0.015, 0.33, 0.015, 0.63), 
    c(0.99, 0.99, 0.99, 0.74, 0.74, 0.49, 0.49), size = 15)

ggsave(filename = "figures/choropleth_grid.png", 
  device = "png", width = 6, height = 5, dpi = 300)
```

@@

Cancer maps are powerful visualisations that summarise complex statistical analyses, however the statistics represented in these maps cannot tell the entire story.
Supplementary graphs and plots are often included to add more depth and information to the map.
@CPISACA suggests additional materials such as 'good tables, graphs, and explanatory text' support understanding and inference derived from maps, ensuring the message communicated will be consistent across a range of viewers.
There are many visualisations used for displays of statistical summaries, these may be dot plots, bar plots, box plots, timesseries plots, cumulative distribution plots, scatter plots, Q-Q plots. 
These additional displays of the cancer distribution can provide alternative views of the cancer statistics, as well as the supporting statistics including error, confidence intervals, distributions, sample or population sizes, standard deviation and other measures.
When presenting cancer maps, @MACM believes the intuition derived from maps must be 'validated by rigorous statistical analyses', the supplementary statistics help for this validation.

```{r supp-grid-create, fig.cap = "Interactive controls of displays in public facing choropleth cancer maps. [a. Globocan 2018: Cancer Today](http://globocan.iarc.fr/Pages/Map.aspx)(Controls for indicator, sex, age groups, continents, and cancer types) , [b. Bowel Cancer Australia Atlas](http://www.bowelcanceratlas.org/) (Controls for indicator, age, year, and geographical areas), [c. The Cancer Atlas ](http://canceratlas.cancer.org/data/#?view=map&metric=INCID_ALL_M) (Controls for predetermined combinations of cancer type or risk factors, mortality or incidence, and gender) [d. Global Cancer Map](http://globalcancermap.com/) (Controls for indicator, and cancer type), [e. United States Cancer Statistics: Data Visualizations](https://gis.cdc.gov/Cancer/USCS/DataViz.html) (Controls for demographics, trends, geography, indicators, risk factors, sex, cancer type, and year)."}
i1 <- png::readPNG('cancer_map_images/interactivity/3.interactivity.png')
i2 <- png::readPNG('cancer_map_images/interactivity/15.interactivity.png')
i3 <- png::readPNG('cancer_map_images/interactivity/4.interactivity.png')
i4 <- png::readPNG('cancer_map_images/interactivity/5.interactivity.png')
i5 <- png::readPNG('cancer_map_images/interactivity/7.interactivity.png')

ggdraw() +
  draw_plot(rasterGrob(i1), 0,  0.13,   .28,  .83) + #a
  draw_plot(rasterGrob(i2), 0.32, 0.56, .68, .4) + #b
  draw_plot(rasterGrob(i3), 0.32, 0.41, .68, .13) + #c
  draw_plot(rasterGrob(i4), 0.32, 0.14, .68, .25) + #d
  draw_plot(rasterGrob(i5), 0, 0, 1, .13) + #e
  draw_plot_label(c("a", "b", "c", "d", "e"), 
    c(0.01, 0.3, 0.3, 0.3, 0.01), 
    c(0.99, 0.99, 0.59, 0.40, 0.19), size = 15)

ggsave(filename = "figures/supp_grid.png", device = "png", 
  width = 6, height = 5, dpi = 300)
```

The interactivity of modern mapping methods enable supplementary information to be incorporated without cluttering the screen.
Tool tip features and drop down menus allow for user exploration, small screens are not a problem with the flexibility provided by zooming and panning [@HTLWM].
This allows relationships between spatial areas and diseases to be explored with sophistication in nontraditional but still 'cognitively accesible' ways [@TNTEA].
The use of these supports were found in a many online diease mapmap sources.
These interactive features provide an opportunity for users to explore additional information to help users understand and interpret the spatial distribution presented, as well as validate, explain or explore the presented statistics and their relationships to each other and/or their underlying spatial distribution. 
This directly enables epidemiologists to explore spatial distributions of cancer statistics.
The interactive features of the publically available maps identified by @roberts2019communication include exploration of geographic hierarchies, population distribution, statistical uncertainty, demographics and socio-economic indicators.
@TNTEA suggested LM plots as a solution to linking cartography and statistical graphics.

@@

### Socio-economic indicators
Socio-economic indicators can explain how the experience of cancer prevalence varies for various members of a society.
These indicators include unemployment rates, poverty rates, remoteness and education levels achieved though, only a few atlases also explored the impact of rurality on cancer rates. 
These rates may also be explored as percentages above or below the mean or median value for the set of spatial areas.
The Human Development Index can be used to understand the socio-economic experience of a community, as can the Income levels as measured by World bank list of economies[@IARC_3]. 
The areas are usually ranked and allocated to quintiles to be presented as categories describing the ranking.

@@

<!-- There are many alternative views of the world-->
<!-- Tobler map projections-->
Choropleths may be considered true topological maps, however, if the land mass displayed covers enough of the globe, there must be a transformation or distortion to display the land in 2D [@HTLWM]. The amount of distortion is related to the distance covered by the landmass displayed @GAMP.
World map projections reflect the frequent perspectives used to view the earth. Choropleth maps will always be distorted if they cover enough of the globe, just like photographs of the globe from space. Choropleth creation requires choosing a map projection that shows a favourable distortion of the geography for presenting the set of spatial information. Selecting a display can prevent misinterpretation of global statistics, as global maps face the challenge of equitable displays of land mass on maps [@RSCW].
If the statistic presented on the map base relies on physical distance and is influenced by the topology there is no transformation needed, beyond choosing a reasonable projection.

@@

<!-- Why transform?-->
<!-- Common variables used to create cartograms, e.g. population, mortality -->
Cartograms provide an alternative visualisation method for statistical and geographical information.
The key difference between a choropleth and a cartogram is the desirable augmentation of the size, shape or distance of geographical areas [@ACTUC]. 
@HTLWM suggests that white lies may be employed to create useful displays and map creators have the ability to draw lines that may distort the geometry and suppress features and it is easy for the average person to disregard the impact of transformations used to create cartograms.
Cartograms may be seen as an extension of map transformations and projections. The favourable distortion is proportional to a value other other than actual earth size area [@NAC].
A disadvantage of the conventional map is that sparsely populated rural areas may be emphasized, whereas the areas representing cities are very small, making interpretation of spatial patterns very difficult.
The distortion of a cartogram accounts for the population density, preventing it from obscuring the spatial patterns [@TAAM]. 
The spatial transformation of map regions relative to the data emphasises the data distribution instead of land size [@CBATCC].
When visualising population statistics @ACTUC considers this equitable representation design 'more socially just', or honest [@NISCC], giving due attention to all members of the population and reducing the visual impact of large areas with small populations [@DMAHP]. 
@HEDP suggests that 'cancer occurs in people, not in geographical areas' and @CTTMB believe that spatial socio-economic data, like cancer rates, are best presented on a cartogram for urban areas as the population map base avoids allocating 'undue prominence' to rural areas. 
@MTMSIH encourage the use of cartograms to highlight small areas and uncover local-level inequalities.

@@

A similar method, where each geographic area is represented using a square, tesselated to create a square grid. This method has been used by FiveThirtyEight, Bloomberg Business, The Guardian, The Washington Post, The New York Times and NPR.
Each area is represented by a square of the same dimensions.
Figure \ref{fig:ggtilegram} shows a square tilegram representing each state using a single square.

@@

Traditionally, epidemiologists as disease mappers would have considered the interactions between the spatial distribution of a diesease, the environment and the people captured by a display.
Recent developments in technology have created an opportunity to migrate atlases from paper map collections to web atlases.
The medium chosen to communicate health data should always be driven by the message to communicate and the needs of the audience.
Where the message is most important, static or animated graphics allow control over the display and interpretation.
It is important for map creators to develop and then test the message being communicated, epidemiologists are able to do this with their expert knowledge.
Static visualisations can help to aid the written results provided by experts, and pairing an explanation with a map or diagram can help to draw attention to a point, group or area of interest.
Static maps do have the caveat of needing to provide enough information for the display to be understandable out of the original context.
Animations do not tolerate or require user input, this makes them useful for directed descriptions of processes, such as directing attention to a particular region or time period [@HTLWM].
Where the needs of the audience is changeable and is the priority, the map creator can allow interactivity for map users to explore.

<!-- Animations -->
@CPISACA provide weather maps as well developed and appropriate examples of spatial displays to communicate to the general public, and encourage those creating disease maps to leverage the success of weather displays.
The movement of a weather system will follow a forecasted path, the animation will communicate this so that all map users can follow the path of the weather system across the geography over a specified period of time.

The move from static, print media displays toward interactivity allow users to develop maps to ask and answer their own questions.
Developmets and adoptions of this technology has allowed map users more control of the message displayed.
Users are able to see the familiar shapes, zoom in to explore densely populated communities, or change the display completely to highlight the population space.
@CPISACA provide three questions that are generally when using a disease map:

@@

A single map display historically created by epidemiologists, may not effectively answer all three at the same time.
This is especially true when areas are small and difficult to identify in a display.
However, with the ability to animate, map creators can specify and answer these questions by directing the attention of a map user to the areas of interest.
Animations can be used to control the message being communicated by capturing and directing attention of users, it will also ensure all users follow the same interpretive path, as their eyes are drawn across the map to the same comparison points.

<!-- Interactivity-->
This oppportunity provides room to move away from the traditional approach of a single map for a single use, into a more explorative territory.
Interactivity allows users to drive their own exploration, controlling a display through variable controls, links, tool tips, and data selection tools [@TGA].
